Deaths at home in NHS Highlands / North Highland LA

### 2021-09-30 highlands_data_and_visualisations.Rmd
completed 2 graphs so far: deaths at home ocer time, and proportion of deaths by place over time - next up is breakdown of causes of death!

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(lemon)  # for showing axis & axis labels in every facet in ggplot
```

@@

### NHS Highland

```{r}
(figure_deaths_at_home_2020_and_2021_nhs_highland <-
  merged_deaths_hb %>%
  filter(ref_area == "Highland") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  mutate(year = factor(year)) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
               names_to = "deaths_type", values_to = "deaths") %>%
  ggplot(aes(x = week_number)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  # geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number, group = deaths_type), show.legend = FALSE) +
  geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, group = deaths_type), show.legend = FALSE, alpha = 0.3) +  # use a fixed alpha value, otherwise the lines are chopped up into rectangular segments and don't look good zoomed in
  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
  scale_x_continuous(breaks = seq(0,60,by=12)) +
  ylim(0, 50) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths",
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
  annotate(geom = "text", x = 15, y = 36, label = "2020", alpha = 0.4) +
  annotate(geom = "text", x = 8, y = 43, label = "2021") +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Weekly deaths (N)",
       title = "NHS Highland",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  annotate(geom = "text", x = 11, y = 5, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom = "curve", x = 11, xend = 15, y = 7, yend = 11, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x = 20, y = 1, label = "Historic mean", colour = "grey20") +
  annotate(geom = "curve", x = 20, xend = 25, y = 3, yend = 20, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  NULL
)

save_output_file(filename = paste0(dir_outputs_highlands,"/deaths_at_home_2020_and_2021_nhs_highlands"), extensions = c(".pdf",".png"), plot = figure_deaths_at_home_2020_and_2021_nhs_highland, width = 18, height = 14, dpi = 300, units = "cm")
```

### Highland LA

```{r}
(figure_deaths_at_home_2020_and_2021_highland_la <-
  merged_deaths_la %>%
  filter(ref_area == "Highland") %>%
  filter(place_of_death %in% c("Home & other non-institution")) %>%
  mutate(year = factor(year)) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
               names_to = "deaths_type", values_to = "deaths") %>%
  ggplot(aes(x = week_number)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  # geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number, group = deaths_type), show.legend = FALSE) +
  geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, group = deaths_type), show.legend = FALSE, alpha = 0.3) +  # use a fixed alpha value, otherwise the lines are chopped up into rectangular segments and don't look good zoomed in
  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
  scale_x_continuous(breaks = seq(0,60,by=12)) +
  ylim(0, 50) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~place_of_death) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths",
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
  annotate(geom = "text", x = 21, y = 31, label = "2020", alpha = 0.4) +
  annotate(geom = "text", x = 10, y = 28, label = "2021") +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Weekly deaths (N)",
       title = "Highland (Local authority)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  annotate(geom = "text", x = 11, y = 5, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom = "curve", x = 9, xend = 10.5, y = 6, yend = 8, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x = 20, y = 1, label = "Historic mean", colour = "grey20") +
  annotate(geom = "curve", x = 20, xend = 25, y = 3, yend = 15, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  NULL
)

save_output_file(filename = paste0(dir_outputs_highlands,"/deaths_at_home_2020_and_2021_highland_la"), extensions = c(".pdf",".png"), plot = figure_deaths_at_home_2020_and_2021_highland_la, width = 18, height = 14, dpi = 300, units = "cm")
```

### Combined

```{r}
# TODO: complete combined graph
(bind_rows(  # combine the two geographical types
  merged_deaths_hb %>%
    filter(ref_area == "Highland") %>%
    filter(place_of_death %in% c("Home & other non-institution")) %>%
    mutate(geography = "Health board"),
  merged_deaths_la %>%
    filter(ref_area == "Highland") %>%
    filter(place_of_death %in% c("Home & other non-institution")) %>%
    mutate(geography = "Local authority")
) %>%
  mutate(year = factor(year)) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
               names_to = "deaths_type", values_to = "deaths") %>%
  ggplot(aes(x = week_number)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  geom_line(aes(y = deaths, colour = deaths_type)) +
  scale_x_continuous(breaks = seq(0,60,by=12)) +
  ylim(0, 50) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(year~geography) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths",
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Weekly deaths (N)",
       title = "Highland (Local authority)",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  annotate(geom = "text", x = 11, y = 5, label = "Historic range (min & max)", colour = "grey20") +
  annotate(geom = "curve", x = 9, xend = 10.5, y = 6, yend = 8, curvature = 0.20, colour = "#4477aa", alpha = 0.5, arrow = arrow(length = unit(0.2, "cm"))) +
  annotate(geom = "text", x = 20, y = 1, label = "Historic mean", colour = "grey20") +
  annotate(geom = "curve", x = 20, xend = 25, y = 3, yend = 15, curvature = 0.20, colour = "grey20", linetype="dashed", arrow = arrow(length = unit(0.2, "cm"))) +
  NULL
)
```

@@

```{r}
# TODO: add NHS highland & LA!
(figure_proportion_deaths_by_place_by_year_highland <-
  bind_rows(
    merged_proportions_of_deaths_by_place %>% mutate(geography = "scotland"),
    proportions_of_deaths_by_place_by_geography
    ) %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  mutate(geography = case_when(
    geography == "hb" ~ "NHS Highland",
    geography == "la" ~ "Highland (Local Authority)",
    tolower(geography) == "scotland" ~ "Scotland",
    TRUE ~ NA_character_
  )) %>%
  filter(
    geography == "Scotland" | (!is.na(ref_area) & ref_area == "Highland")
  ) %>%
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point() +
  geom_line(size = 1) +
  geom_text(data = ~filter(.x, year==2021), aes(x=2021.1, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  scale_x_continuous(breaks = 2015:2021) +
  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
  scale_colour_viridis_d(option="C") +
  facet_wrap(~geography) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(
    x = "Year", y= "Proportion of yearly deaths (%)",
    colour = "",
    subtitle = paste0("* ",only_2021_text),
    caption = paste0(c(source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL
)

save_output_file(filename = paste0(dir_outputs,"/figure_proportion_home_deaths_over_time"), extensions = c(".pdf",".svg",".png"), plot = figure_proportion_home_deaths_over_time, width = 13, height = 10, units = "cm", dpi = 300)
```

>>>>

library(tidyverse)  # for tidy workflow
library(opendatascot)  # importing data from ScotGov open data website
library(phsmethods)  # methods for working with PHS data
library(readxl)  # for reading excel files
# library(SPARQL)  # taken care of by opendatascot
library(lubridate)  # dealing with dates
library(janitor)  # for cleaning column names
library(ISOweek)  # for computing date from ISO week number + year
library(sf)  # for mapping
library(ggrepel)  # for 'mapping'repelling' labels/texst in ggplot
library(patchwork)  # for assembling plots
library(extrafont)  # for working with fonts
library(openxlsx)  # for creating xlsx files
library(knitr)  # for displaying tables in rmd
library(lemon)  # for showing axis & axis labels in every facet in ggplot
library(patchwork)  # for assembling multiple plots
```

@@

```{r}
annotations_deaths_at_home_highland <- tribble(
  ~geography, ~place_of_death, ~x, ~y, ~label,
  "NHS Highland", "Home & other non-institution", 15, 36, 2020,
  "NHS Highland", "Home & other non-institution", 8, 43, 2021,
  "Highland (Local authority)", "Home & other non-institution", 21, 31, 2020,
  "Highland (Local authority)", "Home & other non-institution", 10, 28, 2021,
) %>% mutate(alpha = if_else(label==2021, 1, 0.4)) %>%   # change alpha for 2020
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)

arrows_deaths_at_home_highland <- tribble(
  ~geography, ~geom, ~place_of_death, ~label, ~x, ~y, ~xend, ~yend,
  "NHS Highland", "text", "Home & other non-institution", "Historic range (min & max)", 11, 5, NA, NA,
  "NHS Highland", "curve", "Home & other non-institution", NA, 11, 7, 15, 11,
  "NHS Highland", "text", "Home & other non-institution", "Historic mean", 20, 1, NA, NA,
  "NHS Highland", "curve", "Home & other non-institution", NA, 23, 3, 29, 18
) %>% 
  mutate(place_of_death = factor(place_of_death, levels = c("Hospital", "Home & other non-institution", "Care home")))  # needs to be factor with same levels, otherwise it messes up level order when added as geom_text later!)

(figure_deaths_at_home_2020_and_2021_highland <-
  bind_rows(  # combine the two geographical types
    merged_deaths_hb %>%
      filter(ref_area == "Highland") %>%
      filter(place_of_death %in% c("Home & other non-institution")) %>%
      mutate(geography = "NHS Highland"),
    merged_deaths_la %>%
      filter(ref_area == "Highland") %>%
      filter(place_of_death %in% c("Home & other non-institution")) %>%
      mutate(geography = "Highland (Local authority)")
  ) %>%
  mutate(year = factor(year)) %>%
  pivot_longer(cols = c(deaths_all_causes, deaths_non_covid),
               names_to = "deaths_type", values_to = "deaths") %>%
  ggplot(aes(x = week_number)) +
  geom_ribbon(aes(ymin = deaths_min, ymax = deaths_max), fill = "#4477aa", alpha = 0.5) +
  geom_line(aes(y = deaths_mean), linetype = "dashed", colour="grey20") +
  # geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, alpha = week_number, group = deaths_type), show.legend = FALSE) +
  geom_line(data = . %>% filter(year==2020), aes(y = deaths, colour = deaths_type, group = deaths_type), show.legend = FALSE, alpha = 0.3) +  # use a fixed alpha value, otherwise the lines are chopped up into rectangular segments and don't look good zoomed in
  geom_line(data = . %>% filter(year==2021), aes(y = deaths, colour = deaths_type)) +
  scale_x_continuous(breaks = seq(0,60,by=12)) +
  ylim(0, 50) +  # start scale at 0 for better understanding the scale of deaths
  facet_grid(~geography) +
  scale_colour_manual(values = c("red","blue"),  # custom legend for the two lines
                      labels = c("All deaths",
                                 "Non Covid-19 deaths"),
                      guide = guide_legend(label.hjust = 0.5, override.aes = list(size = 5))
                      ) +
  scale_alpha_continuous(range = c(0.2, 0.6), guide = NULL) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(x = "Week number",
       y = "Weekly deaths (N)",
       title = "Highland: Health Board & Local Authority",
       subtitle = date_range_text,
       caption = paste0(c(death_registrations_text, source_nrs_text), collapse="\n"),
       colour = "Cause of death"
       ) +
  ## annotate historic range
  geom_text(data = annotations_deaths_at_home_highland %>% filter(label==2021), aes(x=x, y=y, label=label), fontface = "bold") +
  geom_text(data = annotations_deaths_at_home_highland %>% filter(label==2020), aes(x=x, y=y, label=label), alpha = 0.6) +
  geom_text(data = arrows_deaths_at_home_highland %>% filter(geom=="text"), aes(x=x,y=y,label=label)) +
  geom_curve(data = arrows_deaths_at_home_highland %>% filter(geom=="curve") %>% slice(1), aes(x=x,y=y,xend=xend,yend=yend), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm")), alpha = 0.5) +
  geom_curve(data = arrows_deaths_at_home_highland %>% filter(geom=="curve") %>% slice(2), aes(x=x,y=y,xend=xend,yend=yend), curvature = 0.20, arrow = arrow(length = unit(0.2, "cm")), linetype = "dashed") +
  NULL
)

save_output_file(filename = paste0(dir_outputs_highlands,"/deaths_at_home_2020_and_2021_highland"), extensions = c(".pdf",".png"), plot = figure_deaths_at_home_2020_and_2021_highland, width = 24, height = 14, dpi = 300, units = "cm")
```

@@

```{r}
(figure_proportion_deaths_by_place_by_year_highland <-
  bind_rows(
    merged_proportions_of_deaths_by_place %>% mutate(geography = "scotland"),
    proportions_of_deaths_by_place_by_geography
    ) %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  mutate(geography = case_when(
    geography == "hb" ~ "NHS Highland",
    geography == "la" ~ "Highland (Local Authority)",
    tolower(geography) == "scotland" ~ "Scotland",
    TRUE ~ NA_character_
  )) %>%
  filter(
    geography == "Scotland" | (!is.na(ref_area) & ref_area == "Highland")
  ) %>%
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point() +
  geom_line(size = 1) +
  geom_text(data = ~filter(.x, year==2021), aes(x=2021.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  scale_x_continuous(breaks = 2015:2021) +
  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
  scale_colour_viridis_d(option="C") +
  facet_wrap(~geography) +
  theme(legend.position = "top", plot.caption = element_text(size = 10, colour = "gray60")) +
  labs(
    x = "Year", y= "Proportion of yearly deaths (%)",
    colour = "",
    title = "Highland Health Board & Local Authority",
    subtitle = paste0("* ",only_2021_text),
    caption = paste0(c(source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL
)

save_output_file(filename = paste0(dir_outputs_highlands,"/figure_proportion_home_deaths_over_time_highland"), extensions = c(".pdf",".png"), plot = figure_proportion_deaths_by_place_by_year_highland, width = 22, height = 10, units = "cm", dpi = 300)
```
